Higher-Order unification for free!

### 2024-04-12 paper/main.tex
add bib
done 4 (can be spellchecked)

\subsubsection{Occur check}
Since compilation moves problematic terms out of the sight of \Ue{},
that procedure can only perform a partial occur check. For example the
unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
but should be disallowed if a $\linkStore$ contains a link like
$\linketaM{}{Y}{\lambda z.X_z}$: we don't know yet if $Y$ will feature
a lambda in head position, but we surely know it contains $X$, hence
$f~Y$ and that fails the occur check.
The procedure \elpiIn{occur-check-links} is in charge of
performing this check that is needed in order to
guarantee \fullref{prop:fidelity}.

@@

First and foremost problematic terms stored in
\linkStore have to be moved back into the game.
Since links are of the form uvar = term (\fullref{inv:linklhs})
and are duplicate free (see dedup-beta dedup-eta),
one can turn a link X = t into an assignment $X \mapsto t$.
This can in general be achieved by unifying X with t. The case
where t is not in \llambda (link beta/llam) is discussed in section xx.

The second step amounts at allocating new variables in the
memory of \Fo{}. In particular some unif problems
such as $F x y = F x z$ requires to allocate a 
variable $G$ so that the assignment
$F_{ab} \mapsto G_a$ can be used to perform
required pruning.

The last step amounts at decompiling each assignment.
Decompiling a term is trivial. An assignment has an
abs node, as in move, can be eliminated by replacing
the bound variable by the actual term in scope.
In order to do this, one needs the \mapStore to be
a bijection. This is the job of \cref{sec:invariant1}.
\todo{polish}

\todo[inline]{dire che però si passa per una subst in cui ste abs
le cambio in lam. Nel codice Coq ci scrivevamo il tipo nella
arity, e quindi sappiamo fare i lambda bene, senza perdita di
informazione. Qui i lam non hanno info, facile. Ma in generale
bisogna spiegare come ci si salva. Ci dormo su: o non generiamo
la subst ma solo il primo termine (la query iniziale) istanziato
(funziona sempre, la prova è quella sopra) oppure bisogna siegare
tutto sto casino e serve un po' di spazio.}

%\input{code/decompile_type}

% \noteE{What is commit-links and complete-mapping?}

% \noindent
% Decompiling an assignment requires to turn abstractions into
% lambdas. For aesthetic purposes we also eta-contract the result
% (not needed since \Fo{} equality can do that)

% \input{code/decompm}

% \noindent
% Finally decompiling a term is trivial, now that we have an extended
% mapping containing all unassigned variables \Ue may have introduced.

% \input{code/decomp}

% \noindent
% Note that we use beta to build fapp nodes when needed (if Ag is empty
% no \elpiIn{fapp} node should appear).

@@

\input{code/unif_fo}

The code given so far still makes no use
of the higher order nature of the ML unif language,
indeed the scope of unif variables generated by the
compiler is always empty, so \Ue{} is first order.

Still, if \foUnifPb is already \wellb, we can set up a proof
that will also work when comp enforces \wellb and  hstep
preserves it, and when terms in \llambda are mapped to ho variables
with a scope.

\begin{lemma}[Compilation round trip] If
  \elpiIn{comp S T [] M [] _ [] _} then \elpiIn{decomp M T S}
\end{lemma}
\begin{proof}[Proof sketch]
trivial if the mapping is a bijection and 
the terms are beta normal.
some discussion about commit maybellam to be done later.
\end{proof}

\begin{lemma}
\Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo above
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as
\Eo on ground terms. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can find this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the \Fo{} terms is only one interesting cases:
\begin{itemize}
\item \elpiIn{fuva X ~\Uo~s}. In this case after \elpiIn{comp} we have
  $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
  $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
\end{itemize}
Since the mapping is a bijection occur check in \Ho{} corresponds to occur
check in \Fo{}.
\end{proof}

\begin{theorem}[Fidelity in \wellb] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold
\end{theorem}
\begin{proof}[Proof sketch]
Since \elpiIn{progress1} is trivial \fstep and \hstep are the same, that is
in this
context where input terms are $\beta\eta$-normal and all is \wellb,
\Ue is equivalent to \Uo.
\end{proof}

@@

A second strategy to deal with problem that are in \notllambda is to make
approximations. This is the case for example of the unification algorithm of Coq
used in its type class solver~\cite{sozeau08}. The approximation consists in
forcing a choice (among the others) when the unification problem is outside
\llambda. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
argument of the two terms is the same, therefore $Y$ is assigned to $X a$. Note
that this is of course an approximation, since $\sigma = \{X \mapsto \lambda x.Y, Y \mapsto
\_\}$ is another valid substitution for the original problem. We stress the 
fact that, again, our unification procedure in the meta language can be 
accommodated for this new behavior: given a \linkbeta, if \lhs is not in \llambda,
then \elpiIn{progress} can try to align the rightmost arguments and unify
the resulting heads.

>>>>

\subsubsection{Occur check}\label{sec:oc}
Since compilation moves problematic terms out of the sight of \Ue{},
that procedure can only perform a partial occur check. For example the
unification problem $X \Ue f~Y$ cannot generate a cyclic substitution alone,
but should be disallowed if a $\linkStore$ contains a link like
$\linketaM{}{Y}{\lambda z.X_z}$: we don't know yet if $Y$ will feature
a lambda in head position, but we surely know it contains $X$, hence
$f~Y$ and that fails the occur check.
The procedure \elpiIn{occur-check-links} is in charge of
performing this check that is needed in order to
guarantee \fullref{prop:fidelity}.

@@

First and foremost problematic terms stored in
\linkStore have to be moved back into the game:
a suspended link must be turned into a valid assignment.
This operation is possible thanks to \fullref{inv:linklhs},
thanks to the fact that no link causes an occur-check~(\ref{sec:oc})
and the fact that \linkStore is duplicate free (\cref{TODO}).

The second step amounts at allocating new variables in the
memory of \Fo{}. This technicality is required because some
higher-order unification may need to prune a variable. For example
$F\appsep x\appsep y = F\appsep x\appsep z$ requires to allocate a 
variable $G$ in order to express the assignment $F_{ab} \mapsto G_a$.

The last step amounts at decompiling each assignment.
Decompiling a term is trivial since \mapStore is a bijection.
The only tricky part concerns the \elpiIn{abs} node. In out simple setting
the \elpiIn{flam} node carries no extra info (other then the function body),
so each \elpiIn{abs} node can be trivially converted to a \elpiIn{flam} one.
In the case of CIC, where lambdas carry the type of the bound variable, one
has to store it somewhere. Note how this piece of information is akin
to the arity of variables, that is CIC's unification variables have a
(function) type, and that type can be used to annotate the lambdas
needed in order to express their assignment.

\begin{lemma}[Compilation round trip]\label{prop:comprt} If
  \elpiIn{compile S T [] M [] _ [] _} then \elpiIn{decompile M T S}
\end{lemma}
% \begin{proof}[Proof sketch]
% Trivial since \mapStore is a bijection and 
% the terms are beta normal.
% some discussion about commit maybellam to be done later.
% \end{proof}

@@

Even if the compiler, so far, makes no use 
of the higher-order nature of the meta language (all variables have
an empty scope so far) we can already show the code of \Ue{}.

\input{code/unif_fo}
\todo{Fix real code calling compile (not comp), rename decompm to decomp}

If \wellb{\foUnifPb} then we can prove that \Uo is a good
``first order'' unification algorithm. Later, when the compiler will
enforce it by \cref{prop:w-enforcing} and the proof will be adjusted.

\begin{lemma}
\Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo above
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as
\Eo. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{} terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case after compilation we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

\begin{theorem}[Fidelity in \wellb] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold
\end{theorem}
\begin{proof}[Proof sketch]
Trivial since \elpiIn{progress1} is a no-op and \fstep and \hstep are the same.
Since all terms are $\beta$-normal and \wellb, \Ue is equivalent to \Uo.
\end{proof}

@@

A second strategy to deal with problem that are in \notllambda is to make
approximations. This is the case for example of the unification algorithm of Coq
used in its type-class solver~\cite{sozeau08}. The approximation consists in
forcing a choice (among the others) when the unification problem is outside
\llambda. For instance, in $X \appsep a \appsep b = Y \appsep b$, the last
argument of the two terms is the same, therefore $Y$ is assigned to $X a$. Note
that this is of course an approximation, since $\sigma = \{X \mapsto \lambda x.Y, Y \mapsto
\_\}$ is another valid substitution for the original problem. We stress the 
fact that, again, our unification procedure in the meta language can be 
accommodated for this new behavior: given a \linkbeta, if \lhs is not in \llambda,
then \elpiIn{progress} can try to align the rightmost arguments and unify
the resulting heads.
